Complete representability for canonical extensions of representable cylindric algebras

### 2021-11-27 cylindric-notes/Text.tex
RCA_n (n < omega) and its canonicity

\begin{lemma} \label{rcachar}
${\bf RCA}_{\omega} = {\bf S} \: \mathfrak{Cm} \mathbb{S} {\bf Ud} \mathcal{F} \mathfrak{ct}_{\omega} = {\bf S} \: \mathfrak{Cm} \mathbb{S} {\bf Ud} \mathcal{W} \mathfrak{ct}_{\omega} = {\bf IGws}_{\omega}$
\end{lemma}

@@

Here we use the following fact related to canonical varieties generated by some class of complex algebras. Let ${\bf K}$ be an elementary class of relational structures, then:
\begin{center}
If ${\bf K}$ is closed under $p$-morphic images, generated subframes, and disjoint unious, \\ then ${\bf S} \mathfrak{Cm} {\bf K}$ is a canonical variety.
\end{center}
One may think of this fact a more abstract version of Fine's theorem which claims that every elementary modal logic is canonical \cite{fine1975some}. This version denotes the same fact, but it is formulated in terms of varieties BAOs generated by complex algebras of some atom structures.

@@

\begin{proof}
By Lemma~\ref{rcachar},
\end{proof}

>>>>

\begin{lemma} \ref{rcachar}
${\bf RCA}_{\omega} = {\bf S} \: \mathfrak{Cm} \mathbb{S} {\bf Ud} \mathcal{F} \mathfrak{ct}_{\omega} = {\bf S} \: \mathfrak{Cm} \mathbb{S} {\bf Ud} \mathcal{W} \mathfrak{ct}_{\omega} = {\bf IGws}_{\omega}$
\end{lemma}

@@

Here we use the following fact related to canonical varieties generated by some class of complex algebras. Let ${\bf K}$ be an elementary class of relational structures, then:
\begin{center}
If ${\bf K}$ is closed under $p$-morphic images, generated subframes, and disjoint unious, \\ then ${\bf S} \mathfrak{Cm} {\bf K}$ is a canonical variety.
\end{center}
One may think of this fact a more abstract version of Fine's theorem which claims that every elementary modal logic is canonical \cite{fine1975some}. This version denotes the same fact, but it is formulated in terms of varieties BAOs generated by complex algebras of some atom structures. We provide a more precise formulation of the fact above.

Let ${\bf K}$ be a class of frames, denote the closure of ${\bf K}$ under ultraproducts as ${\bf Pu} {\bf K}$.

\begin{prop}
Let ${\bf K}$ be a class of frames, then ${\bf Pu}{\bf K} \subseteq \mathbb{H}\mathbb{S}{\bf Ud} {\bf K}$ implies that ${\bf S} \mathfrak{Cm} \mathbb{S}{\bf Ud} {\bf K}$ is a canonical variety.
\end{prop}

@@

\begin{proof}
We have ${\bf RCA}_{\omega} = {\bf S} \mathfrak{Cm} \mathbb{S} {\bf Ud} \mathcal{F} \mathfrak{ct}_{\omega}$.
That's enough to show that ${\bf Pu} \mathcal{F} \mathfrak{ct}_{\omega} \subseteq \mathbb{H}\mathbb{S}{\bf Ud} \mathcal{F} \mathfrak{ct}_{\omega}$.
For that, we need the following claim:
\begin{claim}\label{claim1}
${\bf Pu} \mathcal{F} \mathfrak{ct}_{\omega} \subseteq$
\end{claim}
\end{proof}

\section{Canonicity of ${\bf RCA}_n$ for finite $n$}

In this section we consider classes ${\bf RCA}_n$, where $n < \omega$ is finite.

We provide the complete proof of the following theorem \cite[Theorem 3.4.3]{hirsch2013completions}.
\begin{theorem}\label{finitecanon}
Let $\mathcal{A} \in {\bf CA}_n$, then $\mathcal{A}$ is representable iff $\mathcal{A}^{+}$ is completely representable.
\end{theorem}

For that we need such model theoretic notions as saturation and types, see \cite[Section 6.3]{hodges1993model}.

\begin{defin} Let $\mathcal{M}$ be a first-order structure of a signature $L$ and $S \subseteq \mathcal{M}$. Let $L(S)$ be an extension of $L$ with copies of elements from $S$ as additional constants. We assume that $Cnst(L)$ and $S$ are disjoint.

\begin{enumerate}
\item Let $n < \omega$, an $n$-type over $S$ is a set $\mathcal{T}$ of $L(S)$ formulas $A(\overline{x})$, where $\overline{x}$ is a fixed $n$-tuple of elements from $S$. Notation: $\mathcal{T}(\overline{x})$. A type is an $n$-type for some $n < \omega$.
\item An $n$-type $\mathcal{T}(\overline{x})$ is realised in $\mathcal{M}$, if there exists $\overline{m} \in \mathcal{M}^n$ such that $\mathcal{M} \models A(\overline{m})$ for every $A \in \mathcal{T}(\overline{x})$. $\mathcal{M}$ omits $\mathcal{T}(\overline{x})$, if $\mathcal{T}(\overline{x})$ is not realised in $\mathcal{M}$.
\item $\mathcal{T}(\overline{x})$ is finitely satisfied in $\mathcal{M}$, if every finite subtype $\mathcal{T}_0(\overline{x}) \subseteq \mathcal{T}(\overline{x})$ is realised in $\mathcal{M}$. We can reformulate that as $\mathcal{M} \models \exists \overline{a} \bigwedge \limits_{A \in \mathcal{T}_0} A(\overline{a})$.
\item Let $T$ be a theory, then a type $\mathcal{T}$ over the empty set of constants is $T$-consistent, if there exists a model $\mathcal{M} \models T$ such that $\mathcal{T}$ is finitely satisfied in $\mathcal{M}$.
\item Let $\kappa$ be a cardinal, then $\mathcal{M}$ is $\kappa$-saturated, if for every $S \subseteq \mathcal{M}$ with $|S| < \kappa$ every finitely satisfied $1$-type $\mathcal{T}$ is realised in $\mathcal{M}$.
\end{enumerate}
\end{defin}

By default, a saturated model is an $\omega$-saturated model for us.

The useful facts, they are from \cite{chang1990model} and \cite{hodges1993model}:

\begin{fact} Let $\mathcal{M}$ be an FO-structue and $\kappa$ a cardinal, then:
\begin{enumerate}
\item $\mathcal{M}$ is $\kappa$-saturated, iff every finitely satisfiable $\alpha$-type (an arbitrary $\alpha \leq \kappa$) with fewer than $\kappa$ parameters is realised in  $\mathcal{M}$.
\item If $\mathcal{M}$ is $\kappa$-saturated, then $\mathcal{M}$ is $\lambda$-saturated for every $\lambda < \kappa$.
\item \label{saturation} Every consistent theory has a $\kappa$-saturated model and every model has an elementary $\kappa$-saturated extension.
\item Let $(\mathcal{M}_i)_{i < \omega}$ a family of structures of the (at most) countable signature and $D$ a non-principal ultrafilter over $\omega$, then $\Pi_D \mathcal{M}_i$ is $\omega_1$-saturated.
\end{enumerate}
\end{fact}

\subsection{Proof of Theorem~\ref{finitecanon}}

Let $\mathcal{A} \in {\bf CA}_n$, then if $\mathcal{A}$ is completely representable, then $h$, a complete representation of $\mathcal{A}$, is atomic. That is, $(a_1, \dots, a_n) \in h(1)$, then $(a_1, \dots, a_n) \in h(y)$ for some $y \in \operatorname{At}(\mathcal{A})$.

\begin{defin} \label{theory} Let $\mathcal{A}$ be a cylindric algebra of dimension $n < \omega$.
$L(\mathcal{A})$ is the first-order language that consists of equality plus $n$-ary predicate letters $(R^n_a)_{a \in \mathcal{A}}$. The $L(\mathcal{A})$-theory $T_{\mathcal{A}}$ consists of the following sentences:
\begin{enumerate}
\item $A_+(a,b,c) := \forall x_1, \dots, x_n \: (R_a(x_1, \dots, x_n) \leftrightarrow R_b(x_1, \dots, x_n) \lor R_c(x_1, \dots, x_n))$. Informally, that means $\mathcal{A} \models a = b + c$.
\item $A_{-}(a,b) := \forall x_1, \dots, x_n \: (R_a(x_1, \dots, x_n) \leftrightarrow \neg R_b(x_1, \dots, x_n))$. That is, $\mathcal{A} \models a = - b$.
\item $A_{\neq 0}(a) := \exists x_1, \dots, x_n R_a(x_1, \dots, x_n)$. That is, $\mathcal{A} \models a \neq 0$.
\item $A_{c_i}(a) := \forall x_1, \dots, x_n (R_{c_i a}(x_1, \dots, x_n) \leftrightarrow \exists y_1, \dots y_n (R_a(y_1, \dots, y_n) \land x_i = y_j))$, for $i < n$ and $j < n$ such that $i \neq j$. Informally, $\mathcal{A} \models {c_i} a = 1$.
\item $A_{d_{ij}} := \forall x_1, \dots, x_n (R_{d_{ij}}(x_1, \dots, x_n) \leftrightarrow x_i = x_j)$, for $i, j < n$.
\end{enumerate}
\end{defin}

In fact, we need to show the following implication:
\begin{center}
If $\mathcal{A}$ is representable, then $A^{+}$ is completely representable.
\end{center}

Assume that $\mathcal{A}$ is representable, then the theory $T(\mathcal{A})$ is consistent, then it has an $\omega$-saturated model $\mathcal{M}$ by Fact~\ref{saturation}. We have the following claim:
\begin{claim}
The set $U_{x_1,\dots,x_n} = \{ a \in \mathcal{A} \: | \: \mathcal{M} \models R_a(x_1,\dots,x_n)\}$ is an ultrafilter of $\mathcal{A}$, for $x_1,\dots,x_n \in \mathcal{M}$ with $R_{1}(x_1,\dots,x_n)$.
\end{claim}
Those $U_{x_1,\dots,x_n}$'s allow us to represent atoms of $\mathcal{A}^{+}$.

We define a representation of $\mathcal{A}^{+}$ as a map $h : \mathcal{A}^{+} \to 2^{\mathcal{M}^n}$ such that:
\begin{center}
$h : S \mapsto \{ (x_1, \dots, x_n) \in 1^{\mathcal{M}} \: | \: U_{x_1,\dots,x_n} \in S \}$, for $S \in \operatorname{Spec}(\mathcal{A})$.
\end{center}

\begin{claim} Let $A_1, A_2 \in \operatorname{Spec}(\mathcal{A})$
\begin{enumerate}
\item $h(0^{\mathcal{A}^{+}}) = \emptyset$
\item $h(- A_1) = - h(A_1)$
\item $h(1^{\mathcal{A}^{+}}) = 1^{\mathcal{M}}$
\item If $S \subseteq \operatorname{Spec}(\mathcal{A})$, then $h(\bigcup S) = \bigcup \limits_{U \in S} h(U)$
\end{enumerate}
In particular, $h$ is a Boolean homomorphism.
\end{claim}

\begin{proof}
$ $

\begin{enumerate}
\item $h(0^{\mathcal{A}^{+}}) = h(\emptyset) = \emptyset$.
\item From the definiton of $h$.
\item $h(- A_1) = - h(A_1)$

Let $x_1, \dots, x_n \in 1^{\mathcal{M}}$, then we have:

\begin{center}
$(x_1, \dots, x_n) \in h(- A_1)$ iff $U_{x_1, \dots, x_n} \in - A_1$ iff $U_{x_1, \dots, x_n} \notin A_1$ iff $(x_1, \dots, x_n) \notin h(A_1)$
\end{center}
\item Let $S = \bigcup \limits_{i \in I} S_i$, where $S_i \in \operatorname{Spec}(\mathcal{A})$ for every $i \in I$.
Let $(x_1, \dots, x_n) \in 1^{\mathcal{M}}$, then we have:
\begin{center}
$(x_1, \dots, x_n) \in h(\bigcup \limits_{i \in I} S_i)$ iff $f_{x_1, \dots, x_n} \in \bigcup \limits_{i \in I} S_i$ iff $\exists i \in I \:\: f_{x_1, \dots, x_n} \in S_i$ iff $\exists i \in I \:\: (x_1, \dots, x_n) \in h(S_i)$ iff $(x_1, \dots, x_n) \in \bigcup \limits_{i \in I} S_i$
\end{center}
\end{enumerate}
\end{proof}

\begin{claim}
$h$ is injective.
\end{claim}

\begin{proof}
Let $U \in \operatorname{Spec}(\mathcal{A})$. The first is to show that $h(U)$ is non-empty. The following $n$-type:
\begin{center}
$T(x_1, \dots, x_n) = \{ R_a(x_1, \dots, x_n) \: | \: a \in U \}$
\end{center}
if finitely satisfied in $\mathcal{M}$.

Consider $T_0 = \{ R_{a_1}(x_1, \dots, x_n), \dots, R_{a_k}(x_1, \dots, x_n) \} \subseteq T$. Then $a_1, \dots, a_k \in U$ and $a = a_1 \cdot \dots \cdot a_k \in U$. By the instance of the $A_{\neq 0}(a)$-axiom, we have $\mathcal{M} \models \exists x_1, \dots, x_n R_a(x_1, \dots, x_n)$.
$a \leq a_i$ for $i \leq k$, so we have $\mathcal{M} \models \exists x_1, \dots, x_n R_{a_i}(x_1, \dots, x_n)$ for every $a_i$ with $i \leq k$ by the instance of the $A_{+}(a_i, a, a)$-axiom. That makes every finite subtype of $T$ satisfiable, thus the whole type is finitely satifiable in $\mathcal{M}$.
$\mathcal{M}$ is $\omega$-saturated, then $T$ is realised in $\mathcal{M}$ by some $(x_1, \dots, x_n) \in \mathcal{M}^n$ and, moreover, $\mathcal{M} \models 1(x_1, \dots, x_n)$. As we have already said, $U_{x_1, \dots, x_n}$ is an ultrafilter, but $U_{x_1, \dots, x_n} \subseteq U$, thus $U = U_{x_1, \dots, x_n}$, so $(x_1, \dots, x_n) \in h(U)$.

That makes $h$ one-to-one.
\end{proof}

\begin{claim}
$ $

\begin{enumerate}
\item $h(c_i{^{\mathcal{A}^{+}}} U) = C_i (h(U))$
\item $h(d_{ij}^{\mathcal{A}^{+}}) = D_{ij} \subseteq \operatorname{Spec}(\mathcal{A})$
\end{enumerate}
\end{claim}

\begin{proof}
$ $

\begin{enumerate}
\item Let $\overline{x} = (x_1, \dots, x_n) \in \mathcal{M}^{n}$ and $S \subseteq \operatorname{Spec}(\mathcal{A})$. Assume $(x_1, \dots, x_n) \in h(c_i{^{\mathcal{A}^{+}}} S)$.

Let us show that $\overline{x} \in C_i (h(S))$, that is, there exists $\overline{y} = (y_1, \dots, y_n) \in h(S)$ such that $\overline{x} \equiv_{i} \overline{y}$.

Then $\mathcal{M} \models 1(x_1, \dots, x_n)$ and $U_{x_1, \dots x_n} \in c_i{^{\mathcal{A}^{+}}} S$. But $\mathcal{A}^{+}$ is the complex algebra of the ultrafilter frame $\mathcal{F}_{\mathcal{A}}$.
Then we have:
\begin{center}
$c_i{^{\mathcal{A}^{+}}} S = \{ U_1 \in \operatorname{Spec}(\mathcal{A}) \: | \: \exists U' \in S  \: U_1 R_{i} U' \}$
\end{center}
Then there must be an ultrafilter $U' \in S$ such that $U_{x_1, \dots x_n} R_{i} U'$, that is, $c_i a \in U_{x_1, \dots x_n}$ whenever $a \in U'$.
Hence $\mathcal{M} \models R_{c_i}(x_1, \dots x_n)$. By the $A_{c_i}(a)$-axiom, we have
\begin{center}
$\mathcal{M} \models \exists z_1, \dots, z_n (R_a(z_1, \dots, z_n) \land x_i = z_j)$ for $i < n$ and $j < n$ such that $i \neq j$.
\end{center}
Consider the following $n$-type with free variables $z_1, \dots, z_n$ and parameters $x_1, \dots, x_n \in \mathcal{M}$:
\begin{center}
$T(z_1, \dots, z_n) = \{ R_a(z_1, \dots, z_n) \land x_i = z_j \: | \: i < n, j < n, i \neq j, a \in U' \}$.
\end{center}
Let us show that $T(z_1, \dots, z_n)$ is finitely satifsiable in $\mathcal{M}$.
Consider a finite subset of $T$, say $T_0 = \{ R_{b_k}(z_1, \dots, z_n) \land x_i = y_j \: | \: i < n, j < n, i \neq j, b_k \in U', k < \omega \}$.
We put $p = p_1 \cdot \dots \cdot p_k$ and $p \in U'$ since $U'$ is a filter. Then we have:
\begin{center}
$\mathcal{M} \models \exists z_1, \dots, z_n (R_b(z_1, \dots, z_n) \land x_i = z_j)$ for $i < n$ and $j < n$ such that $i \neq j$
\end{center}
Thus, we have, as required:
\begin{center}
$\mathcal{M} \models \exists z_1, \dots, z_n \bigwedge \limits_{i = 1}^{k} (R_{b_k}(z_1, \dots, z_n) \land x_i = z_j)$ for $i < n$ and $j < n$ such that $i \neq j$.
\end{center}
As above, using $\omega$-saturation, we conclude that $T$ is realised in $\mathcal{M}$ at an $n$-tuple $(y_1, \dots, y_n) = \overline{y}$.
Then we have:
\begin{center}
$\mathcal{M} \models 1(\overline{y})$, $\overline{x} \equiv_i \overline{y}$, $U_{\overline{y}} \supseteq U'$
\end{center}
Then $U_{\overline{y}} = U'$, then $\overline{y} \in h(S)$. Then $\overline{x} \in C_i (h (S))$.

Suppose for the converse, $\overline{x} = (x_1, \dots, x_n) \in C_i (h(S))$. We need $\overline{x} \in h(c_i(S))$.
Then there exists $\overline{y} = (y_1, \dots, y_n)$ such that $\overline{x} \equiv_i \overline{y}$ and $\overline{y} \in h(S)$. Then there exists an ultrafilter $U_{y_1, \dots, y_n} \in S$. Let us show that $\mathcal{M} \models 1(x_1, \dots, x_n)$ and $U_{x_1, \dots, x_n} \in c_i U_{y_1, \dots, y_n}$.
Let $a \in U_{y_1, \dots, y_n}$. Then we have $\mathcal{M} \models R_a(y_1, \models, y_n)$. By the $A_{c_i}(a)$ axiom, we have $\mathcal{M} \models R_{c_i a}(x_1, \dots, x_n)$. Then $\mathcal{M} \models 1(x_1, \dots, x_n)$ and $c_i a \in U_{x_1, \dots, x_n}$, thus, $\overline{x} \in h(c_i(S))$.
\item Let us show that $h$ preserves cylindrifications.

Let $(x_1, \dots, x_n) \in \mathcal{M}^n$. Then $(x_1, \dots, x_n) \in D_{ij}$ iff $\mathcal{M} \models 1(x_1, \dots, x_n)$ and $x_i = x_j$ iff $U_{x_1, \dots, x_n} \in d_{ij}^{\mathcal{A}^{+}} = \{ U \in \operatorname{Spec}(\mathcal{A}) \: | \: d_ij \in U \}$ iff $\mathcal{M} \models d_{ij}^{\mathcal{M}}(x_1, \dots, x_n)$.
\end{enumerate}
\end{proof}
